---
title: "Day 1: Introduction to the CCR, and to R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Day 1: Introduction to the CCR, and to R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(RPG520)
```

# Connecting to the CCR

# Essential R

Variables

```{r}
1 + 2
```

```{r}
pie <- "apple"
pie
```

```{r}
pi
```

Vectors

```{r}
1:3
```

```{r}
id <- 1:3
month <- c("Jan", "Feb", "Mar")
sex <- factor(
    c("Male", "Female", "Male"),
    levels = c("Female", "Male")
)
```

```{r}
id > 1
id[id > 1]
```

Special values -- `NA`

```{r}
age <- c(27, NA, 32)
age > 27
age[age > 27]
```

Functions

```{r}
sqrt(id)
```

The `data.frame`

```{r}
df <- data.frame(
    id = id,
    month = month,
    sex = sex
)
```

```{r}
df[df$sex == "Male",]
```

Packages

- CRAN -- more than 20,000 packages (!) <https://CRAN.R-project.org>
  - Use `install.packages()` to install CRAN packages
- Bioconductor -- 2266 packages for high-througput genomic data
  analysis <https://bioconductor.org>
  - First install [BiocManager][] from CRAN, then use
    `BiocManager::install()` to install Bioconductor packages.

[BiocManager]: https://CRAN.R-project.org/package=BiocManager

# 'Tidy' data

```{r, message = FALSE}
library(readr)
library(dplyr)
```

```{r, include = FALSE}
all_file <- system.file(package = "RPG520", "extdata", "ALL.csv")
all <- read_csv(all_file)
```

```{r, eval = FALSE}
all_file <- file.choose()
all <- read_csv(all_file)
```

```{r}
all
```

```{r}
glimpse(all)
```

# Exercise
