---
title: "Day 1: Introduction to the CCR, and to R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Day 1: Introduction to the CCR, and to R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##"
)
```

```{r setup, message = FALSE}
library(RPG520)
```

# Connecting to the CCR

# Introduction to *R*

## Essential *R*

A simple calculator

```{r}
1 + 1
```

'Vectors' as building blocks

```{r}
c(1, 2, 3)
c("January", "February", "March")
c(TRUE, FALSE)
```

Variables, missing values and 'factors'

```{r}
age <- c(27, NA, 32, 29)
gender <- factor(
    c("Female", "Male", "Non-binary", NA),
    levels = c("Female", "Male", "Non-binary")
)
```

Data structures to coordinate related vectors -- the `data.frame`

```{r}
df <- data.frame(
    age = c(27, NA, 32, 29),
    gender = gender
)
df
```

Key operations on `data.frame`

- `df[1:3, c("gender", "age")]` -- subset on rows and columns
- `df[["age"]]`, `df$age` -- select columns

Functions

```{r}
rnorm(5)        # 5 random normal deviates
x <- rnorm(100) # 100 random normal deviates
hist(x)         # histogram, approximately normal
plot(density(x)) # a little more sophisticated?
```

'Vectorized' operations, e.g., element-wise addition without an
explicit 'for' loop. The following creates a vector `y` by
element-wise addition of the 100 elements of vector `x` with 100
random normal deviates.

```{r}
y <- x + rnorm(100)
```

`x` and `y` are related, so place them as columns in a
`data.frame()`. There are 100 rows in the `data.frame()`, `head()`
shows the first 6 rows.

```{r}
df <- data.frame(x, y)
nrow(df)
head(df)
```

There is a statistical relationship between `y` and `x`, specifically,
`y` is equal to `x` plus some random noise. Visualize this using
`plot()`.

```{r, fig.asp = 1}
plot(y ~ x, df)
```

It looks like a linear relationship; can we demonstrate this
statistically? Use the `lm()` function to fit a *l*inear *m*odel to
the relationship between `y` and `x`:

```{r}
fit <- lm(y ~ x, df)
fit           # an R 'object' containing information about the
              # regression of y on x
summary(fit)  # statistical summary of linear regression
```

What about adding the regression line to the scatter plot? Do this by
plotting the points, and then adding `abline()` to add the straight
line implied by the linear model fit.

```{r, fig.asp = 1}
plot(y ~ x, df)
abline(lm(y ~ x, df))
```

### A little more advanced...

Write your own functions

```{r}
hello <- function(who) {
    paste("hello", who, "with", nchar(who), "letters in your name")
}
hello("Martin")
```

Iterate, usually with `lapply()` although `for()` is available

```{r}
names <- c("Martin", "Thomas")
lapply(names, hello)
```

## Packages

Packages extend functionality of base _R_. Can be part of the 'base'
distribution (`lm()` is from the 'stats' package)...

```{r}
stats::lm(y ~ x)
```

... or a package contributed by users to the Comprehensive R Archive
Network ([CRAN][]), or to [Bioconductor][] or other repositories.

- CRAN -- more than 20,000 packages (!) <https://CRAN.R-project.org>
  - Use `install.packages()` to install CRAN packages
- Bioconductor -- 2266 packages for high-througput genomic data
  analysis <https://bioconductor.org>
  - First install [BiocManager][] from CRAN, then use
    `BiocManager::install()` to install Bioconductor packages.

[CRAN]: https://CRAN.R-project.org
[Bioconductor]: https://bioconductor.org
[BiocManager]: https://CRAN.R-project.org/package=BiocManager

A particularly common suite of packages is the 'tidyverse'.  To use
the [dplyr][] contributed CRAN package, load the package

```{r, message = FALSE}
library(dplyr)
```

A dplyr 'tibble' is like a 'data.frame', but more user-friendly

```{r}
tbl <- tibble(
    x = rnorm(100),
    y = x + rnorm(100)
)

tbl # e.g., only displays the first 10 rows
```

The [ggplot2][] package is often used for visualizing data

```{r, message = FALSE}
library(ggplot2)
```

[ggplot2][] uses a 'grammar of graphics' to assemble a plot in layers

```{r, fig.asp = 1}
ggplot(tbl) +                    # use 'tbl' as a source of data for the plot
    aes(x, y) +                  # use columns 'x' and 'y' for axes
    geom_point() +               # add x and y coordinates as points
    geom_smooth(method = "lm")   # add a regression line
```

[dplyr]: https://CRAN.R-project.org/package=dplyr
[ggplot2]: https://CRAN.R-project.org/package=ggplot2

## How to Learning *R*

Help

- Every function has a help page, e.g.,

    ```{r, eval = FALSE}
    ?data.frame
    ```

    - Sections with 'Description', 'Usage', 'Arguments' (often with
      defaults), 'Value', 'Examples'
    - Can be quite difficult to follow, but are usually quite helpful.

Vignettes

- Many packages have 'vignettes' that illustrate how the package can
  be used.

    ```{r, eval = FALSE}
    vignette(package = "dplyr")
    ## Vignettes in package 'dplyr':
    ##
    ## colwise                 Column-wise operations (source, html)
    ## base                    dplyr <-> base R (source, html)
    ## grouping                Grouped data (source, html)
    ## dplyr                   Introduction to dplyr (source, html)
    ## programming             Programming with dplyr (source, html)
    ## rowwise                 Row-wise operations (source, html)
    ## two-table               Two-table verbs (source, html)
    ## in-packages             Using dplyr in packages (source, html)
    ## window-functions        Window functions (source, html)

    vignette(package = "dplyr", "dplyr")
    ```

Books

- [Hands-on Programming with R][hopr] introduces *R* programming.
  [The Very Basics][hopr-basics] provides an elementary starting
  point.
- [R for Data Science][r4ds] is a book-length web spanning with both
  introductory and more advanced topics. One of the authors, Hadley
  Wickham, is the author of the [dplyr][] and [ggplot2][] packages,
  among others.

Web resources

- Google can be a great place to find general *R* resource, e.g., 'how
  to do survival analysis in *R*'
- [StackOverflow][] is useful for trouble-shooting particular problems
  -- 'I try to fit a linear model with `lm(mgp ~ cyl, mtcars)` but *R*
  says ..., what am I doing wrong?'.
- ChatGPT & friends can also be useful, especially if asking
  well-focused questions or, e.g., asked to 'explain the following *R*
  code' to help understand code someone else has written. I have
  almost always disagreed with the specific code ChatGPT generates;
  sometimes it simply does not work (e.g., use the package 'X', but
  'X' is not a package), but more dangerous are the situations where
  code is syntactically correct but provides the wrong answer. Use,
  but with extreme care!

[hopr]: https://rstudio-education.github.io/hopr/
[hopr-basics]: https://rstudio-education.github.io/hopr/basics.html
[r4ds]: https://r4ds.hadley.nz/
[StackOverflow]: https://stackoverflow.com/questions/tagged/r

# Grading for this week

The task for this week is to write an *R* script that performs data
management and statistical analysis of a data set of your choice --
essentially reproducing selected steps in the work that we will do on
Wednesday.

You will provide me (on the CCR server) with a file 'your_name.R'. I
will run the file in a new *R* session using the command
`source('your_name.R', echo = TRUE)`. This will read and evaluate each
*R* command in the file.

Grading will be out of 10, with the following thresholds

- 7 / 10 if `source('your_name.R', echo = TRUE)` works without error.
- 8 / 10 for scripts that implement more extensive analyses, or that
  present interesting or complicated data.
- up to 10 / 10 for work that goes beyond the material we cover this
  week, e.g., using 'Rmarkdown' to present your work, performing
  additional (and appropriate) data manipulation steps using dplyr or
  other packages, additional visualizations or statstical analyses.

Note that the 7 / 10 score could be obtained by copying & pasting (a
subset of) the commands from Wednesday's lab into
'your_name.R'. Please feel free to contact or speak with me if you
have problems or ideas.

# Session information {.unnumbered}

For reproducibility, I record the software versions used to create
this document

```{r}
sessionInfo()
```
