---
title: "Appendix"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Appendix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##"
)
```

# Installation

## *R*

Install *R* for your operating system from the Comprehensive *R*
Archive Network [CRAN][]. The box at the top of the page has links for
Linux, macOS, and Windows.

On Windows, see the link [install *R* for the first
time][CRAN-windows]. The link will download an installer that can be
double-clicked. Install as a regular user, not as a system
administrator.

On macOS, use the 'R-4.3.2-arm64.pkg' installer for newer (M1 / M2
macOS) and 'R-4.3.2-x86_64.pkg' for Intel macs.

Linux systems can of course be more complicated; follow the links for
your operating system.

[CRAN]: http://CRAN.R-project.org/
[CRAN-windows]: https://CRAN.R-project.org/bin/windows/base/

## *RStudio*

It is possible to use *R* without using *RStudio* (this is how I
usually use *R*, for instance), but many people use the open-source
version of *RStudio*. Download the open-source version of [RStudio
Desktop][RStudio] from Posit, the company that produces *RStudio*.

[RStudio]: https://posit.co/download/rstudio-desktop/

# Packages

If you are installing *R* on your own laptop for the purposes of this
course, you'll need to install the packages that we use.  *R* on the
CCR has many packages pre-installed. But there are more than 20,000
packages in CRAN and Bioconductor, so at some point you will need a
package that is not installed.

Follow this general approach. Note that packages only need to be
installed *once*; once installed, they can then be used in separate
*R* sessions using `library()` or the `pkg::function()` notation.

## CRAN package installation

Start by creating a vector of packages to install. For this workshop,
we used the following packages.

```{r}
pkgs <- c(
    "readr", "dplyr", "tidyr", "ggplot2",
    "plotly", "survival", "ggsurvfit"
)
```

Use `setdiff()` (remove from the vector in the first argument all
elements that occur in the vector in the second argument) so that only
packages that are not currently installed remain.

```{r}
pkgs <- setdiff(pkgs, rownames(installed.packages()))
```

Finally, using the function `install.packages()` to install the
required packages from a central CRAN repository.

```{r, eval = FALSE}
install.packages(pkgs, repos = "https://CRAN.R-project.org")
```

## Bioconductor packages

[Bioconductor][] packages require a slightly different installation
procedure. Make sure that you have the [BiocManager][] package
installed from CRAN

```{r, eval = FALSE}
pkgs <- "BiocManager"
pkgs <- setdiff(pkgs, rownames(installed.packages()))
install.packages(pkgs, repos = "https://CRAN.R-project.org")
```

Then install [Bioconductor][] (and CRAN) packages as needed

```{r, eval = FALSE}
pkgs <- c(
    "cellxgenedp", # Bioconductor package
    "Seurat"       # CRAN package
)
pkgs <- setdiff(pkgs, rownames(installed.packages()))
BiocManager::install(pkgs)
```

## Updating packages

`install.packages()` and `BiocManager::install()` report packages that
are out-of-date because a newer version is available. Usually it is a
good idea to be using the most recent version available. There are two
situations when you might wish to continue using older packages.

- A tight deadline (e.g., thesis defense, paper submission). Updating
  packages can sometimes introduce changes that break existing code,
  and it can be time consuming to track these down.

- Reproducibility. Packages sometimes introduce changes that result in
  (hopefully slightly) different outcomes, perhaps because the
  packages adopts a newer algorithm. This might not be desireable,
  e.g., when a paper has been submitted and the reviewer says 'this is
  excellent expect ...' it requires a minor addition to the orginal
  analysis, but when you try and reproduce your original analysis it
  is no longer possible because of changes in the underlying software.

  There are a number of strategies for taking a 'snapshot' of software
  versions used in a particular analysis (e.g.,, the [renv][]
  package), and if you become serious about using *R* it is very
  valuable to explore these resources *before* you need to reproduce
  them.

## FAQ: binary or source packages?

On Windows and macOS, one might sometimes see a message that indicates
a 'binary' package is available, but the 'source' package is more
recent. Almost always the best response is to stay with the 'binary'
package -- the more-recent 'source' package will likely be made
available in a short time frame (within a week), and if not indicates
that the package is difficult to install even by the wizards at CRAN
-- mere mortals like you and I will have to become very proficient to
install Windows or macOS packages from source.

[Bioconductor]: https://bioconductor.org/packages
[renv]: https://rstudio.github.io/renv/index.html

# Data used in this workshop

Datasets used in this workshop are available on the CCR. They can also
be accessed on the internet. The URL for each dataset is the
`base_url` pasted before the file name used in this document. Thus the
ALL dataset is available at

```{r}
base_url <-
    "https://raw.githubusercontent.com/mtmorgan/RPG520/main/inst/extdata/"
ALL_url <- paste0(base_url, "ALL.csv")
```

It can be read directly into *R*

```{r, eval = FALSE}
all <- readr::read_csv(ALL_url)
```

Or downloaded to disk for easy re-use

```{r, eval = FALSE}
download.file(ALL_url, "ALL.csv")
```
